\vspace{-2.0cm}The effect of nitrogen application time on infection and development of the Wheat Streak Mosaic Virus (WSMV)

### 2015-10-21 leslieandkenny.Rnw
done with count issues

<<data,echo=FALSE>>=
# Read and set up the data
#for leslie
setwd("~/Documents/ConsultingFall2015/Fall510")
fert2 <- read.csv("fert2.csv")

@@

\section{Introduction}
\noindent

\section{Statistical Experience and Assistance Needed}
\noindent Nar B. Ranabhat is a PhD candidate in the Department of Land Resources and Environmental Science. He is currently working on several projects regarding the spread and development of the Wheat Streak Mosaic Virus (WSMV) in Winter Wheat. Nar has taken Statistics $511$ and $512$ here at Montana State University, and he is currently sitting in on Mixed Models. Nar has asked for our help in building a model, fitting it in R, and interpreting the results.

@@

\begin{figure}[h!]
\begin{center}
\caption{Illustration of separation.}
<<separation,echo=FALSE,out.width='0.6\\linewidth',fig.width=6>>=
curve(exp(x)/(1+exp(x)), from = -10, to = 10, ylim = c(0, 1),
      xlab = expression(logit(p[i])), ylab = expression(p[i]))
segments(x0 = -10, x1 = log(1/59), y0 = 1/60, lty = 2, lwd = 2)
segments(x0 = log(1/59), y0 = 0, y1 = 1/60, lty = 2, lwd = 2)
@
\end{center}
\end{figure}

\noindent The following sections present examples of issues that could
arise while analyzing Nar's data. The examples all use simulated datasets.

@@

\noindent We noticed that infection counts of 0 were recorded in 67 of the
180 plots in Nar's dataset. This is a large fraction of the data; if the
distribution of these zeros is related to the variables in the model then
it may be impossible to estimate coefficients for factor levels where
infection rates are nearly zero. \\

\noindent We recommend sorting the data and creating a heatmap to look
for patterns of zeros. Figure 2 shows an example using simulated data.
The most serious problems will come from the zeros, so we chose a contrasting
color (black) for plots with infection counts of 0.\\

\noindent If a treatment has many zeros across all varieties or a variety
has many zeros, Nar should consider whether he has reason to believe
that treatment or variety has qualitative differences from the others
that would cause it to have a much lower infection rate. These data would
be considered as coming from a different population than the data with
fewer zeros. \emph{If and only if it is justified}, Nar could omit all
of the data for this treatment or variety when fitting the model. The scope
of inference for this model must then be limited to varieties and
treatments that have a nonzero infection rate. A separate analysis would
be needed for the more resistant population. \\

@@

@@

\noindent The model is
$$
y_i\sim\text{overdispersed Binomial}(30, p_i, \omega),
$$
$$
\mathrm{logit}(p_i)=\mu+\alpha_{b[i]}+\beta_{v[i]}+\gamma_{t[i]}+\delta_{j[i]}
+(\alpha\beta)_{b[i],v[i]}+(\beta\gamma)_{v[i],t[i]}+(\beta\delta)_{v[i],j[i]}
+(\gamma\delta)_{t[i],j[i]}+(\beta\gamma\delta)_{v[i],t[i],j[i]}
$$
where
\begin{itemize}
\item \(p_i\) is the probability of infection in the \(i\)th plot,
\item \(\mathrm{logit}(p_i)=\log\left(\dfrac{p_i}{1-p_i}\right)\) is the
natural logarithm of the odds of infection, and
\item \(b[i]\), \(v[i]\), \(t[i]\), and \(j[i]\) index the block,
variety, nitrogen application timing, and inoculation status of
observation \(i\),
\item \(\omega\) is an overdispersion parameter, so that the variance
of the response in plot \(i\) is \(30\omega p_i(1-p_i)\).
\end{itemize}

@@

\noindent The model is
$$
y_i\sim\mathrm{Binomial}(30, p_i),
$$
\begin{align*}
\mathrm{logit}(p_i)&=\mu+a_{b[i]}+(ab)_{b[i],v[i]}\\
&\quad+\beta_{v[i]}+\gamma_{t[i]}+\delta_{j[i]}\\
&\quad+(\alpha\beta)_{b[i],v[i]}+(\beta\gamma)_{v[i],t[i]}
+(\beta\delta)_{v[i],j[i]}+(\gamma\delta)_{t[i],j[i]}\\
&\quad+(\beta\gamma\delta)_{v[i],t[i],j[i]}
\end{align*}
where
\begin{itemize}
\item \(p_i\) is the probability of infection in the \(i\)th plot,
\item \(\mathrm{logit}(p_i)=\log\left(\dfrac{p_i}{1-p_i}\right)\) is the
natural logarithm of the odds of infection, and
\item \(b[i]\), \(v[i]\), \(t[i]\), and \(j[i]\) index the block,
variety, nitrogen application timing, and inoculation status of
observation \(i\),
\item \(a_b\sim\mathrm{N}(0,\sigma^2_a)\) and
\((ab)_{b,v}\sim\mathrm{N}(0,\sigma^2_{ab})\) are random adjustments for
each block and each variety within each block.
\end{itemize}

>>>>

<<data,echo=FALSE>>=
# Read and set up the data
#for leslie
#setwd("~/Documents/ConsultingFall2015/Fall510")
fert2 <- read.csv("fert2.csv")

@@

\section{Introduction}

\noindent Nar B. Ranabhat is a PhD candidate in the Department of Land Resources and Environmental Science. He is currently working on several projects regarding the spread and development of the Wheat Streak Mosaic Virus (WSMV) in Winter Wheat. Nar has taken Statistics $511$ and $512$ here at Montana State University, and he is currently sitting in on Mixed Models. Nar has asked for our help in building a model, fitting it in R, and interpreting the results.

@@

The response variable is the count of infected leaves out of the 30 leaves
collected in each plot. The standard approach to analyzing count data is
to use a Binomial distribution and model the probability \(p_i\) that a leaf
in a given plot \(i\) will get infected. This modeling occurs on the scale
of the log-odds, \(\mathrm{logit}(p_i)=\log\left(\frac{p_i}{1-p_i}\right)\). \\

\noindent One serious problem that can occur when working with counts is
the issue of separation. This is when one variable or a combination of factors
is perfectly associated with the response. For example, if a certain
treatment results in an infection count of zero, then knowing that a leaf
recieved that treatment determines that the leaf will not get infected,
regardless of other variables. This causes difficulty in estimating the
effect of that treatment on the infection rate, and results a large standard
error because the effect cannot be precisely determined. \\

\noindent Figure 1 provides a visual illustration of the separation problem.
The left panel shows a small probability on the vertical axis and its
corresponding logit value on the horizontal axis. The right panel zooms
in on the lower left corner. The dotted lines show how a small amount
of uncertainty on the probability scale translates to a relatively larger
amount of uncertainty on the logit scale. \\

\begin{figure}[h!]
\begin{center}
\caption{Illustration of separation.}
<<separation,echo=FALSE>>=
par(mfrow = c(1, 2), mar = c(4, 4, 2, 2))
curve(exp(x)/(1+exp(x)), from = -10, to = 10, ylim = c(0, 1),
      xlab = expression(logit(p[i])), ylab = expression(p[i]))
segments(x0 = -10, x1 = log(1/59), y0 = 1/60, lty = 2, lwd = 2)
segments(x0 = log(1/59), y0 = 0, y1 = 1/60, lty = 2, lwd = 2)

curve(exp(x)/(1+exp(x)), from = -10, to = -2, ylim = c(0, 0.4),
      xlab = expression(logit(p[i])), ylab = expression(p[i]))
segments(x0 = -10, x1 = log(1/59), y0 = 1/60, lty = 2, lwd = 2)
segments(x0 = log(1/59), y0 = 0, y1 = 1/60, lty = 2, lwd = 2)
segments(x0 = -10, x1 = log(1/39), y0 = 1/40, lty = 3, lwd = 1)
segments(x0 = log(1/39), y0 = 0, y1 = 1/40, lty = 3, lwd = 1)
segments(x0 = -10, x1 = log(1/119), y0 = 1/120, lty = 3, lwd = 1)
segments(x0 = log(1/119), y0 = 0, y1 = 1/120, lty = 3, lwd = 1)
@
\end{center}
\end{figure}

\noindent The problem gets worse if the infected count is 0. In that
case, the best estimate of the infection probability is 0, and
\(\mathrm{logit}(0)=-\infty\). This will cause computational problems
because the software will try to improve the model fit by decreasing
the effect estimate toward \(-\infty\). The software may fail to find a
solution, and if it does report a solution, the standard error will be
large because the solution cannot be found precisely. \\

\noindent The following sections present examples of issues that could
arise while analyzing Nar's data, and our recommendations on dealing with
them. The examples all use simulated datasets.

@@

\noindent A graphical analysis is also a good way to see if separation
is present in the data. In Nar's dataset, 67 of the 180 counts are
0. This is a large fraction of the data, so it is very important to look
for possible separation. We recommend sorting the data and creating a
heatmap to look for patterns of zeros. Figure 2 shows an example using
simulated data. The most serious problems will come from the zeros, so we
chose a contrasting color (black) for plots with infection counts of 0. \\

@@

\noindent If a treatment has many zeros across all varieties or a variety
has many zeros, Nar should consider whether he has reason to believe
that treatment or variety has qualitative differences from the others
that would cause it to have a much lower infection rate. These data would
be considered as coming from a different population than the data with
fewer zeros. \emph{If and only if it is justified}, Nar could omit all
of the data for this treatment or variety when fitting the model. The scope
of inference for this model must then be limited to varieties and
treatments that have a nonzero infection rate. \\

\noindent The graphic could be presented as evidence that the omitted
data does not fit the same model as the data with higher infection counts.
A separate analysis would be needed if additional information is desired
about the more resistant population. \\

@@

\noindent The model is
$$
y_i\sim\text{overdispersed Binomial}(30, p_i, \omega),
$$
\begin{align*}
\mathrm{logit}(p_i)&=\mu+\alpha_{b[i]}+\beta_{v[i]}
+\gamma_{t[i]}+\delta_{j[i]}\\
&\quad+(\alpha\beta)_{b[i],v[i]}+(\beta\gamma)_{v[i],t[i]}
+(\beta\delta)_{v[i],j[i]}+(\gamma\delta)_{t[i],j[i]}
+(\beta\gamma\delta)_{v[i],t[i],j[i]}
\end{align*}
where
\begin{itemize}
\item \(p_i\) is the probability of infection in the \(i\)th plot,
\item \(\mathrm{logit}(p_i)=\log\left(\dfrac{p_i}{1-p_i}\right)\) is the
natural logarithm of the odds of infection, and
\item \(b[i]\), \(v[i]\), \(t[i]\), and \(j[i]\) index the block,
variety, nitrogen application timing, and inoculation status of
observation \(i\),
\item \(\omega\) is an overdispersion parameter, so that the variance
of the response in plot \(i\) is \(30\omega p_i(1-p_i)\).
\end{itemize}

@@

\noindent The model is
$$
y_i\sim\mathrm{Binomial}(30, p_i),
$$
\begin{align*}
\mathrm{logit}(p_i)&=\mu+a_{b[i]}+(ab)_{b[i],v[i]}
+\beta_{v[i]}+\gamma_{t[i]}+\delta_{j[i]}\\
&\quad+(\alpha\beta)_{b[i],v[i]}+(\beta\gamma)_{v[i],t[i]}
+(\beta\delta)_{v[i],j[i]}+(\gamma\delta)_{t[i],j[i]}
+(\beta\gamma\delta)_{v[i],t[i],j[i]}
\end{align*}
where
\begin{itemize}
\item \(p_i\) is the probability of infection in the \(i\)th plot,
\item \(\mathrm{logit}(p_i)=\log\left(\dfrac{p_i}{1-p_i}\right)\) is the
natural logarithm of the odds of infection, and
\item \(b[i]\), \(v[i]\), \(t[i]\), and \(j[i]\) index the block,
variety, nitrogen application timing, and inoculation status of
observation \(i\),
\item \(a_b\sim\mathrm{N}(0,\sigma^2_a)\) and
\((ab)_{b,v}\sim\mathrm{N}(0,\sigma^2_{ab})\) are random adjustments for
each block and each variety within each block.
\end{itemize}
